---
title: "Treemendous: Example"
author: "Andrea Paz, Felix Specker, Dan Maynard"
date: "2023-07-17"
output:
  pdf_document: 
    keep_tex: true
fontsize: 11pt
---


## Package Installation

```{r install, eval=FALSE}
library(devtools)
install_github("speckerf/treemendous")
```

## Species List Preparation

All functions of \textit{Treemendous} require the species name to be split into two columns, \verb|Genus| and \verb|Species|, with the former being capitalized. Assume you have two species, \textit{Acer platanoides} and \textit{Fagus sylvatica}, you can create the input \verb|tibble| by calling: 


```{r input, message=FALSE}
### Species list preparation
library(tidyverse)
species <- c('Acer platanoides', 'Fagus sylvatica')
input <- species %>%
  tibble::as_tibble_col(column_name = 'binomial') %>%
  tidyr::separate(col = 'binomial', into = c('Genus', 'Species'))
input
```

Other useful functions for creating the input \verb|tibble| include:

```{r, eval=FALSE}
readr::read_csv('path') # import data
dplyr::select(Genus, Species) # select columns
dplyr::distinct(Genus, Species) # remove duplicate binomials
dplyr::rename('Genus' = 'old_genus_name',
                'Species' = 'old_species_name') # rename columns
dplyr::mutate(Genus = stringr::str_to_title(Genus)) # capitalize Genus
dplyr::mutate(Species = stringr::str_remove(Species, ".*?\\s")) # remove everything before first space
tidyr::drop_na(c('Genus', 'Species')) # remove rows with NA's
dplyr::arrange(Genus, Species) # sort names
dplyr::bind_rows(x, y) # concatenate two tibble's

```

## FIA: Standardize species names from the U.S. Forest Inventory and Analysis program. 

Along with the package comes an example dataset \verb|fia| with $2171$ different tree species names. Assume that we want to standardize these species names according to a certain backbone (use the backbone argument). The function \verb|summarize_output()| can be used to get a summary of the process. 

```{r}
library(treemendous)
```

```{r, message=FALSE}
result <- fia %>% matching(backbone = 'BGCI')
summarize_output(result)
```

\noindent From $2171$ species names in total, we were able to match $1822$ according to the backbone \verb|BGCI|, with $1779$ names matching exactly, and $43$ species names matching using fuzzy- and suffix-matching. Besides information about the matching process, the output contains the old names (prefix \verb|Orig.|) as well as the matched names (prefix \verb|Matched.|) as follows:

```{r}
result %>% 
  dplyr::slice_head(n=3) %>%
  dplyr::select(1:5)
```

\noindent We can further increase the number of matched species by using the functions \verb|matching()| followed by \verb|enforce_matching()|. Here, we specify the backbone \verb|BGCI|. 

```{r, message=FALSE}
result <- fia %>% 
  matching(backbone = 'BGCI') %>% 
  enforce_matching(backbone = 'BGCI')
result %>% summarize_output()
```

\noindent Now, we are able to match $2097$ species names in total, with $275$ species being matched via \verb|enforce_matching()|. Note that the number of matched distinct species names is lower with $2044$, because several input species were matched to the same species in the target database \verb|BGCI|. 

\noindent Note that if we choose a different backbone than \verb|BGCI|, then species can matched names that are not accepted (synonyms), we can further resolve synonyms after matching the species names with the function \verb|resolve_synonyms()|. Now, the output contains additionally the accepted species names (prefix \verb|Accepted.|), as well as a column \verb|Accepted.Backbone|, which states according to which backbone the synonym was resolved.

```{r, message=FALSE}
result <- fia %>% 
  matching('WFO') %>% 
  resolve_synonyms('WFO')
```

```{r}
result %>% 
  dplyr::slice_head(n=3) %>% 
  dplyr::select(dplyr::matches('Orig|Matched|Accepted'), -'matched')
```
\noindent Note that a warning message is produced "Please consider calling highlight_flags() to investigate potential ambiguities upon resolving synonyms to accepted names". Potential ambiguities could have been resolved in your dataset and it is suggested to use \verb|highlight_flags()| to know more and decide if you want to check them manually. The \verb|highlight_flags()| function should be used separately from the others as it will only return species that have some flag and not the full dataset. Note also that each entry can have multiple flags:

```{r, message=FALSE}
flags <- result %>% highlight_flags('WFO')
flags %>% 
  dplyr::slice_head(n=3) %>% 
  dplyr::select(dplyr::matches('Acc|ambiguity|link'))
```
We can see the full breakdown of these flags as follows:
```{r, message=FALSE}
flags %>% dplyr::select(dplyr::contains("WFO")) %>% dplyr::summarize_all(.funs = sum)
```
The bulk of these flags denotes an \verb|infraspecific_ambiguity|, which can generally be ignored, provided that the user did not manually truncate any trinomials to binomials for input. The $37$ \verb|infraspecific_link| flags are likewise typically not problematic, as these simply highlight when the input binomial differs from the output binomial via a trinomial link at some point in the graph. The remaining $142$ \verb|authorship_ambiguity| are the most problematic, as these indicate taxa that have multiple conflicting matches. These should be manually explored and used with caution.

\noindent Instead of using a single backbone, the user can decide to use any subset of the backbones \verb|c('BGCI', 'WFO', 'WCVP', 'GBIF')| or use all of them by simply calling \verb|matching()| without any argument. While \verb|matching()| considers all backbones being equally important, the function \verb|sequential_matching()| can be used to call \verb|matching()| for individual backbones sequentially. For every species, the matched backbone is provided in the column \verb|Matched.Backbone|. 

```{r, eval=F, message=FALSE}
result <- fia %>% 
  sequential_matching(sequential_backbones = c('BGCI', 'WFO', 'WCVP'))
```
\noindent Remember that \verb|matching()| and \verb|sequential_matching()| match any species in the database and thus can provide matches to synonyms rather than accepted species. To get only accepted species returned use \verb|resolve_synonyms()| after the matching function.

## Translate species names between two databases.

Oftentimes, researches require integrating multi-modal data from
different sources for their analyses. Here, we demonstrate the use of the 
function \verb|translate_trees()|, which allows a user directly translate names from an input database to a target database. 
First, we resolve both databases individually according to the single backbone (WFO) and compare the resolved names. Then, we use translate\_trees to translate the input species names into the target names.

```{r}
input <- tibble::tibble(
  Genus = c('Aria', 'Ardisia', 'Malus'),
  Species = c('umbellata', 'japonica', 'sylvestris')
)
target <- tibble::tibble(
  Genus = c('Sorbus', 'Ardisia', 'Malus'),
  Species = c('umbellata', 'montana', 'orientalis')
)
```

\small
```{r, message=FALSE}
input %>%
  matching(backbone = 'WFO') %>%
  resolve_synonyms('WFO') %>%
  dplyr::select(1:6)
```

```{r, message=FALSE}
target %>%
  matching(backbone = 'WFO') %>%
  resolve_synonyms('WFO') %>%
  dplyr::select(1:6)
```
\normalsize
Resolving both sets individually leads to a mismatch - _Malus orientalis_ and _Malus sylvestris_ were resolved to two different names. Now let's see whether translate\_trees can be used to match all three species: 
\small
```{r, message=FALSE}
translate_trees(df = input, target = target) %>% 
  dplyr::select(1:4) 
```
\normalsize
Essentially, all three species names can be translated from the input set to the target set. Incorporating the knowledge of the desired target names, the function leverages the information about synonym-accepted relations in the three backbones WFO, WCVP and GBIF and is able to translate _Malus sylvestris_ into _Malus orientalis_. 

